R for Clinical Study Reports and Submission

### 2021-10-27 slides/workshop-slides.Rmd
add slides for ch 9 - 14

---
title: "R for Clinical Study Reports and Submission"
subtitle: "R/Pharma Conference 2021"
author: "Yilong Zhang, Nan Xiao, Keaven Anderson"
date: "October, 2021"
output:
  mkdocs::mk_ioslides_presentation:
    mk_logo: "invent"
    mk_col_title: "white"
    mk_col_section: "teal"
    mk_class: "public"
    widescreen: true
---

@@

> The tool, process and format may be different in different organization/company.
> We only provide a recommended way to address them.

@@

- [Contributors](https://r4csr.org/) of the training materials.
  + Please consider to submit issues or PR in the repo.

- [R consortium](https://www.r-consortium.org/projects/isc-working-groups)
  + R validation hub
  + Submission working group
  + R Tables for Regulatory Submission (RTRS) working group
  + Please consider to join and contribute to a working group.

- R / Pharma organizers
  + It is a fun and productive annual gathering. 
  + Please consider to share story and use cases to enlarge the community.

@@

```{r, echo = TRUE, eval = FALSE}
install.packages(c(
 "tidyverse",        # Data Manipulation 
 "r2rtf",            # TLF generation
 "pkglite"           # eCTD Submission 
))
```

@@

- Training Material: https://r4csr.org/
- Demo project: 
  + Analysis project: https://github.com/elong0527/esubdemo
  + eCTD submission package: https://github.com/elong0527/ectddemo
- R Consortium FDA submission pilot: 
  + https://github.com/RConsortium/submissions-pilot1

- During the workshop, we will use `esubdemo` project.
  + `esubdemo` is shared in RStudio Cloud (https://rstudio.cloud/) 
  + Project link will be shared in chat
  + Post questions in group chat.

@@

## ICH E3 guidance

The [ICH E3: strcuture and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) provide guidance to assist sponsors in the development of a CSR.

@@

- The dataset structure is following [CDISC Analysis Data Model (ADaM)](https://www.cdisc.org/standards/foundational/adam).

- Source data: https://github.com/elong0527/r4csr/tree/master/adam_data

@@

- `r2rtf`: an R package to create production ready tables and figures in RTF format.

@@

Different organization can have different table standards

@@

Before creating an RTF table we need to:

@@

- Let's refer examples in the [Section 2.4.2](https://r4csr.org/overview.html#r2rtf)

@@

`r2rtf` provides simple "verb" functions that correspond to each component of a table, 
  to help you translate data frame to table in RTF file.

@@

https://r4csr.org/disposition.html

@@

https://r4csr.org/population.html

@@

https://r4csr.org/baseline-characteristics.html

@@

https://r4csr.org/efficacy.html

@@

https://r4csr.org/aesummary.html

@@

https://r4csr.org/specific-ae.html

@@

In this part, let’s consider how to organize a clinical trial project as an A&R leader.

@@

https://r4csr.org/folder.html

@@

https://r4csr.org/manage.html

@@

>>>>

---
title: "R for Clinical Study Reports and Submission"
subtitle: "R/Pharma Conference 2021"
author: "Yilong Zhang, Nan Xiao, Keaven Anderson"
date: "October 2021"
output:
  mkdocs::mk_ioslides_presentation:
    mk_logo: "invent"
    mk_col_title: "white"
    mk_col_section: "teal"
    mk_class: "public"
    widescreen: true
---

@@

> The toolchain, process, and formats may be different in different organizations.
> We only provide a recommended way to address them.

@@

- [Contributors](https://r4csr.org/) of the training materials.
  + Please consider submitting issues or PR in the repo.

- [R Consortium](https://www.r-consortium.org/projects/isc-working-groups)
  + R validation hub
  + Submission working group
  + R Tables for Regulatory Submission (RTRS) working group
  + Please consider to join and contribute to a working group.

- R / Pharma organizers
  + It is a fun and productive annual gathering.
  + Please consider sharing stories and use cases to expand the community.

@@

```{r, echo = TRUE, eval = FALSE}
install.packages(c(
  "tidyverse",      # Data manipulation
  "r2rtf",          # TLF generation
  "pkglite"         # eCTD submission
))
```

@@

- Training Material: <https://r4csr.org/>
- Demo project: 
  + Analysis project: <https://github.com/elong0527/esubdemo>
  + eCTD submission package: <https://github.com/elong0527/ectddemo>
- R Consortium FDA submission pilot: 
  + <https://github.com/RConsortium/submissions-pilot1>

- During the workshop, we will use `esubdemo` project.
  + `esubdemo` is shared in RStudio Cloud (<https://rstudio.cloud/>)
  + Project link will be shared in chat
  + Post questions in group chat.

@@

## ICH E3 Guidance

The [ICH E3: structure and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) provide guidance to assist sponsors in the development of a CSR.

@@

- The dataset structure follows the [CDISC Analysis Data Model (ADaM)](https://www.cdisc.org/standards/foundational/adam).

- Source data: <https://github.com/elong0527/r4csr/tree/master/adam_data>

@@

- `r2rtf`: an R package to create production-ready tables and figures in RTF format.

@@

Different organizations can have different table standards

@@

Before creating an RTF table, we need to:

@@

- Let's refer to examples in the [Section 2.4.2](https://r4csr.org/overview.html#r2rtf)

@@

`r2rtf` provides simple "verb" functions that correspond to each component of a table, 
  to help you translate data frame to tables in RTF file.

@@

<https://r4csr.org/disposition.html>

@@

<https://r4csr.org/population.html>

@@

<https://r4csr.org/baseline-characteristics.html>

@@

<https://r4csr.org/efficacy.html>

@@

<https://r4csr.org/aesummary.html>

@@

<https://r4csr.org/specific-ae.html>

@@

In this part, let's consider how to organize a clinical trial project as an A&R lead.

@@

Our primary focus is creating a standard R package structure to organize the
project, with 4 goals in mind:

- Consistency
- Reproducibility
- Automation
- Compliance

<https://r4csr.org/folder.html>

@@

- Setting up for success
  - Teamwork
  - Architecture
  - Boundaries
  - Community

- The SDLC
  - Planning
  - Development
  - Validation
  - Operation

<https://r4csr.org/manage.html>

@@

<https://r4csr.org/overview-2.html>

## Submission Package

We will discuss strategies to prepare proprietary R packages and analysis code
into proper formats for submission:

- The whole game
- Practical considerations
- Prepare R packages using `pkglite`
- Prepare analysis programs
- Update ADRG and ARM

<https://r4csr.org/submission-package.html>

## Running Environment

We will also give recommendations to make the R code
running environment reproducible for dry run tests and reviews
using `cleanslate`.

<https://r4csr.org/running-environment.html>

# pkglite

## pkglite: Compact Package Representations

- To provide a **tool** for packing and restoring R packages as plaintext
assets that are easy to store, transfer, and review.
- To provide a **grammar** for specifying the file packing scope that is
functional, precise, and extendable.
- To provide a **standard** for exchanging the packed asset that is
unambiguous, human-friendly, and machine-readable.

## Pipe-Friendly Workflow

```{r, eval=FALSE, echo=TRUE}
library("pkglite")

"/path/to/pkg/" %>%
  collate(file_ectd()) %>%
  pack()

pack(
  "/path/to/pkg1/" %>% collate(file_ectd()),
  "/path/to/pkg2/" %>% collate(file_ectd()),
  output = "/path/to/pkglite.txt"
)

"/path/to/pkglite.txt" %>%
  unpack(output = "/path/to/output/")
```

## File Specifications + File Collections

File specifications offer flexibility and brevity in specifying the files to include.

| File specification type | Functions                                                                           |
|-------------------------|-------------------------------------------------------------------------------------|
| Manual discovery        | `file_spec()`                                                                       |
| Automatic discovery     | `file_auto()`                                                                       |
| Common patterns         | `file_root_core()`, `file_r()`, `file_man()`, `file_src()`, `file_vignettes()`, ... |
| Default sets            | `file_default()`, `file_ectd()`                                                     |

<br>

File collections contain the evaluation results of file specifications for packing.

```{r, eval=FALSE, echo=TRUE}
library("pkglite")

"/path/to/pkg" %>%
  collate(file_root_core(), file_r(), file_auto("inst/"))
```

## pkglite.txt

pkglite.txt follows the standard Debian Control File (DCF) format used by
Debian, R, and RStudio IDE, to be both machine-readable and human-readable.

```
# Generated by pkglite: do not edit by hand
# Use pkglite::unpack() to restore the packages

Package: pkg1
File: DESCRIPTION
Format: text
Content:
  Package: pkg1
  Type: Package
  Title: Example Package One
  Version: 0.1.0
```
